---
title: "lab18"
author: "Max Wang"
toc: True
format: pdf
---
## Background

Pertussis (also known as whooping cough) is a common lung disease caused by the bacteria Bordetella pertussis. This disease can spread to all individuals however when infecting infants below the age of 1, it becomes a potentially deadly disease. To combat this, a vaccine was developed in 1946 which protected against the disease. Recently however, pertussis cases have increased resulting in a heightened risk for infant death. This lab seeks to investigate cdc data to gain insight on the potential sources of this increase.

## Loading PDB Data

From the cdc [website](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html) we can scrape this data and store it as a data frame. 

>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(datapasta)
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L,
    2024L,
    2025L
  ), cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063,
    22538,
    21996
  )
)
```
> Q. Make a plot of `year` vs `cases`

```{r}
library(ggplot2)

plot <- ggplot(cdc, aes(year, cases)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Cases of Pertusis")
plot
```

>Q2 Adding some major milestones. The introduction of wP vacines in 1946, the swich to the newer aP vacine in 1996, and the COVID years in 2020.

```{r}
plot <- ggplot(cdc, aes(year, cases)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Cases of Pertusis") +
  geom_vline(xintercept = 1946, lty = 2, col = "blue") +
  geom_vline(xintercept = 1996, lty = 2, col = "red") +
  geom_vline(xintercept = 2020, lty = 2, col = "green")
plot
```

There were high case numbers before 1940 then cases dropped dramatically and remained low between the 1970s and the 2000. After the early 2000s, the case numbers suddenly rose again into the 2010s.

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine and a delay, there was a slow increase in Pertussis cases. Furthermore considering case data in other countries that switched suggest a similar offset between the aP vaccine switch and new cases. This suggests that the aP vaccine seems to have a degrading effect over time.

##Investigation with the CMI-PB project

The Computational Models of Immunity - Pertussis Boost (CMI-PB) project aims to provide information about how large scale implementation of the aP vaccine to provide information about how the aP vaccines effectiveness wanes over time compared to the wP vaccine.

This data is avaliable online in a JSON format via their [website](https://www.cmi-pb.org/). To do so we can use the function `read_jason()` from the jsonlite package.

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5_1/subject", simplifyVector = T)
head(subject)
```
>Q4 how many wP and aP indidivuals are in the table:

```{r}
table(subject$infancy_vac)
```
There are 87 aP and 85 wP individuals '

>Q5 what is the biological sex breakdown of the data set

```{r}
table(subject$biological_sex)
```
There are 112 females and 60 males 

>Q6 In terms of race and gender, is this representative of the US population 

```{r}
table(subject$race, subject$biological_sex)
```
It is not representative of the population however the data is still usable due to the size and relative scope. 

Reading more data tables:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = T)
head(specimen)
```

```{r}
antibody <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = T)
head(antibody)
```

For analysis, first merge the data using the the `*_join` function in the **dplyr** library.

```{r}
library(dplyr)
meta <- inner_join(subject,specimen)
head(meta)
```

```{r}
abdata <- inner_join(antibody, meta)
head(abdata)
```

>Q What antibody isotypes are measured for these patients

```{r}
table(abdata$isotype)
```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$antigen)
```
>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```
The most recent dataset has significantly more observations than the previous two datasets but its still smaller than the first 2020 dataset. 



Looking at the IgG isotype, plotting the MFI_normalized for all antigens. 

```{r}
igg <- abdata %>% 
  filter(isotype == "IgG")
```

```{r}
ggplot(igg, aes(MFI_normalised, antigen)) +
  geom_boxplot()
```

>Q Is there a difference between aP vs wP values

```{r}
ggplot(igg, aes(MFI_normalised, antigen, col = infancy_vac)) +
  geom_boxplot() + 
  facet_wrap(~visit)
```


## Focus on PT (Pertussis toxin) antigen

```{r}
pt_igg_2021 <- igg %>% 
  filter(antigen == "PT") %>%
  filter(dataset == "2021_dataset")
head(pt_igg_2021)
```

Plotting the PT antigen experssion over time

```{r}
ggplot(pt_igg_2021, aes(planned_day_relative_to_boost, MFI_normalised, col = infancy_vac, group = subject_id)) +
  geom_point() +
  geom_line()+
  geom_vline(xintercept = 14, lty = 2, col = "black")
```
Adding polish and a trend line

```{r, warning=FALSE}
ggplot(pt_igg_2021, aes(planned_day_relative_to_boost, MFI_normalised, col = infancy_vac, group = subject_id)) +
  geom_point(alpha = 0.7) +
  geom_line(alpha = 0.3)+
  geom_vline(xintercept = 14, lty = 2, col = "black") +
  geom_vline(xintercept = 7, lty = 2, col = "black") +
  geom_smooth(data = subset(pt_igg_2021, planned_day_relative_to_boost <= 7), aes(group = infancy_vac), se = FALSE) +
  geom_smooth(data = subset(pt_igg_2021, planned_day_relative_to_boost >= 7), aes(group = infancy_vac), se = FALSE)
```


