---
title: "lab08"
author: "Max Wang"
format: pdf
toc: True
---

## Background

Using the techniques learned throughout the class, we will be analyzing real breast cancer biopsy data through PCA.

## Preparing the Data

load the data sets into R variables. 
```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
head(wisc.df)
```

removing diagnosis from the data frame and storing it in a separate variable diagnosis 

```{r}
wisc.data <- wisc.df[,-1]
head(wisc.data)

diagnosis <- wisc.df$diagnosis

nrow(wisc.data) #Q1

table(diagnosis) #Q2

length(grep("mean", colnames(wisc.data), fixed = T)) #Q3
```
>Q1. How many observatiosn are in this data set? 

569

>Q2. How many observations have a malignant diagnosis? 

212

>Q3. How many variables/features are suffixed with _mean? 

10

## Applying PCA to the Data

Before using PCA we first check to ensure the variances between each variable is consistent 

```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)
```

Using PCA on the data set

```{r}
wisc.pr <- prcomp(wisc.data, scale = T)
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

44.27% of the variance.

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 principal components.

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 prinipal components 

## Visualizing PCA Results

First trying to visualize results using the `biplot()` function

```{r}
biplot(wisc.pr)
```
>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

Since everything is scatterd messily and lablels are overlapping, there isnt anything that stands out

Using ggplot2 to plot PC1 and PC2

```{r}
library(ggplot2)
pr_plot_1_2 <- ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(aes(col = diagnosis))
pr_plot_1_2
```

Further plotting PC1 and PC3

```{r}
pr_plot_1_3 <- ggplot(wisc.pr$x) +
  aes(PC1, PC3) +
  geom_point(aes(col = diagnosis))
pr_plot_1_3
```

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

There are a clear clustering of points that represent the two doctor diagnoses.


## Variance Explained

Identifying the scree point for principal component variance by plotting the variance by each successive component.

Creating a vector that contains the proportion of total variance in each PC.  

```{r}
pr.var <- wisc.pr$sdev^2
pve <- pr.var/sum(pr.var)
head(pve)
```
Plotting the variation across each added principal component

```{r}
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")

```

From the plot above, there is a clear scree point at the third principal component.

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

There are not, the other features have a lower contribution to the PC1 vector. 

```{r}
wisc.pr$rotation[,1]
max(abs(wisc.pr$rotation[,1]))
```



## Heirarchical Clustering



First scale the data and assign distances between each point.

```{r}
scaled.data <- scale(wisc.data)
data.dist <- dist(scaled.data)
```

Create a hierarchical clustering model using `hclust()`

```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```

Creating a cluster deprogram using the `hclust()` result

```{r}
plot(wisc.hclust)
abline(h = 19, col = "red", lty = 2)
```

>Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

roughly h = 19

Using `cutree()` to create clusters of the data and compares to the clinically diagnosed clusters.

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, h = 19)
table(wisc.hclust.clusters, diagnosis)
```
Using different `method` arguments in `hclust()`

```{r}
plot(hclust(data.dist, method = "ward.D2"))
plot(hclust(data.dist, method = "single"))
plot(hclust(data.dist, method = "average"))
```

>Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning. 

The `ward.D2` method gives the cleanest results since it creates a plot with clear branches that show the clustering results. Furthermore with `ward.D2` it avoids small branches like in `complete`.

## Combining Heirarchical Clustering with PCA

Using `hclust()` on the PCA data using enough components to account for 90% of the data. transformed the PCA data into distances from each point. 

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:3]), method = "ward.D2")
plot(wisc.pr.hclust)
abline(h = 70, col = "red", lty = 2)
```

Using `cutree()` to assign groups and comparing to the known diagnoses.

```{r}
grps <- cutree(wisc.pr.hclust, k = 2)
table(grps, diagnosis)
```
Cluster 1: 179 "M" diagnoses
Cluster 2: 333 "B" diagnoses 
179 true positives
24 false positives
333 true negatives
33 false negatives <- the worst case, try to minimize

Specificity Tp/(TP + FN)
```{r}
179/(179+33)
```

Sensitivity TN/(TN + FP)
```{r}
333/(333+24)
```
>Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

Overall within the two groups that were assigned, the specificity of the groups was 0.84 and the sensitivity was 0.93 so `hclust()` model separates "M" and "B" well. 

```{r}
table(wisc.hclust.clusters, diagnosis)
```

Ignoring clusters 2 and 4:
TP: 165
FP: 12
TN: 343
FN: 40

Specificity Tp/(TP + FN)
```{r}
165/(165 + 40)
```


Sensitivity TN/(TN + FP)

```{r}
343/(343 + 12)
```


>Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? 

Comparatively, the clustering off of the raw data results in a more messy result, creating 4 separate clusters with 2 clusters without clear distinctions. Numerically, the PCA model creates groups with higher specificity while the raw data creates models with higher sensitivity. Overall, with the messier groups using the raw data, the PCA groupings are more effective. 



Adding groups to the PC1 vs PC2 plot.

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col = grps)
```


## Predicting Diagnoses using PCA results

Using the groups from the PCA results, we can predict the potential diagnoses by projecting them on the PCA space.

Importing new samples

```{r}
new <- read.csv("new_samples.csv")
```

Using the `predict()` function, we can use the previous PCA ouput to assign the new samples PC scores using the PCs from the PCA. 

```{r}
npc = predict(wisc.pr, newdata = new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

>Q16. Which of these new patients should we prioritize for follow up based on your results?

We should follow up on patient 1 because patient 1's tumor is within the group that has a high likelihood of being malignant.

