---
title: "Lab05"
author: "Max Wang"
format: pdf
toc: true
---

## Background
There are lot's of ways to make plots in R. These include so-called "base R" (like the `plot()` function) and add on packages like **ggplot2**

Let's make the same plot with these two graphics systems. We can use the inbuilt `cars` data set:


```{r}
head(cars)
```

With "base R" we can make graphs using the `plot()` function 

```{r}
plot(cars)
```

Now using ggplot, first install the package using `install.packages(ggplot2)`

> **N.B.** do not place install.packages in a code chunk otherwise it repeatedly installs every time the document is rendered. 

after installation of the library, the library needs to be loaded with a call to the function `library()`

```{r}
#install.packages("ggplot2")  ggplot installation code 
library(ggplot2)
```

ggplot objects need at least 3 things:

1. the `data` to use to be graphed 
2. the `aes` for aesthetics that map the data of the graph
3. the `geom` for the plot type 

```{r}
  ggplot(data = cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  labs(x = "Speed (mph)", y = "Distance (ft)", title = "Stopping Distance by Speed") + 
  theme_bw()
```
## Gene Expression Plot

Using data on gene expression with and without GLP-1 inhibitors we can plot a graph to highlight significant genes that are up or down regulated. 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

version 1 plot - simple dot plot comparing gene expression with and without treatment.

```{r}
ggplot(genes) +
  aes(x = Condition1, y = Condition2) +
  geom_point(col = "blue")
```

Using the `State` value for each gene, we can color the graph based on the state within `aes()` and specify the colors using `scale_color_manual()`

```{r}
ggplot(genes) +
  aes(x = Condition1, y = Condition2, col = State) +
  geom_point() +
  scale_color_manual(values = c("blue", "gray", "red")) +
  labs(x = "Control (No Drug)", 
       y = "Drug Treatment", 
       title = "Gene Regulation of GLP-1 Inhibitor Treatment") +
  theme_classic()
```

## Further Exploration with gapminder 

Using the `gapminder` data set we look to explore further capabilities of ggplot2

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"
gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many rows (total entries) does the data set have 

```{r}
nrow(gapminder)
```

> how many different continents are in the data set

```{r}
unique(gapminder$continent)
```

Version 1 plot: `gdpPercap` vs `lifeExp` 

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point() +
  labs(x = "GDP per Capita (usd)", y = "Life Expectency", title = "Life Expectancy vs GDP per Capita by Continent")
```

Creating separate plots for each continent using the "faceting" feature in ggplot:

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point() +
  facet_wrap(~continent) +
  labs(x = "GDP per Capita (usd)", y = "Life Expectency", title = "Life Expectancy vs GDP per Capita by Continent")
```

## Using the dplyr library

Adding another package `dplyr` that allows for the usage of the `filter()` function to select data points from the data set 

```{r}
library(dplyr)
```

```{r}
filter(gapminder, year == 2007, country == "United States")
```

```{r}
input <- filter(gapminder, year == 2007 | year == 1977)
```

```{r}
ggplot(input) +
  aes(x = gdpPercap, y = lifeExp, col = continent, size = pop) +
  geom_point() +
  facet_wrap(~year) +
  theme_classic()
```

