---
title: "lab12"
author: "Max Wang"
format: pdf
toc: True
---
## Identifying Genetic Varients

>Q1: What are those 4 candidate SNPs?

rs12936231, rs8067378, rs9303277, and rs7216389

>Q2: What three genes do these variants overlap or effect?

ZPBP2, IKZF3, and GSDMB

>Q3: What is the location of rs8067378 and what are the different alleles for rs8067378?

rs8067378 is on chromasome 17:39894595-39895595, and it has the alleles A/G

>Q4: Name at least 3 downstream genes for rs8067378?

GSDMB, ORMDL3, and LRRC3C

>Q5: What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are
homozygous for the asthma associated SNP (G|G)?

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (1).csv")
table(mxl$Genotype..forward.strand.)/nrow(mxl)
```
14.06% of the sample of people with Mexican Ancentry were homoxygous G|G for the asthma SNP. 


Similar example with CHB (han Chinese in bejing)
```{r}
mxl <- read.csv("373515-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
table(mxl$Genotype..forward.strand.)/nrow(mxl)
```


>Q6. Back on the ENSEMBLE page, use the “search for a sample” field above to find the
particular sample HG00109. This is a male from the GBR population group. What is the
genotype for this sample?

This sample has the genotype G|G

## Introduction to RNA Seq Analysis 

Using Galaxy, upload sequencing data and use built in tools to analyze the RNA seq data. 

>Q7: How many sequences are there in the first file? What is
the file size and format of the data? 

There are 3,863 sequences in the first file and it is 741.9 KB total in fastqsanger format. 

**there are two files because sanger produces a forward and a reverse read**

**FASTQ format starts with a @ instead of a > like fasta. comments are still in (). The next line is the actual sequence followed by a new line an +. After the + is the quality score. quality score determines how confident the nt is. Quality is based of numbering translated to ascii, A = 64. Its diff for each format but overall all same idea**

```
@title (comment)
ATGCAT 
+
AA/EEE (score)
```
>Q8: What is the GC content and sequence length of the second fastq file?

running the fastQC program to test quality of reads. 

The second fastaq file has a gc content of 54% and a length of 50-75 nt

>Q9: How about per base sequence quality? Does any base have a mean quality score below
20?

There were non with a mean quality score below 20

## Mapping RNA seq data to a genome 

Effectively using a known genome as a guide to align each fragment and map them to different genes in the genome. Based on the number of sequence reads we can then determine expression levels. 


>Q10: Where are most the accepted hits located?

Most of the accepted hits are within the range chr17:38,150,862-38,154,477.

>Q11: Following Q10, is there any interesting gene around that area?

The hits are within the exons on the gene PSMD3.

Using the tool cufflinks and an already anotated sequence of the human genome, we can find sequence expression by coudnting the RNA seq hits. 

>Q12: Cufflinks again produces multiple output files that you can inspect from your right-handside galaxy history. From the “gene expression” output, what is the FPKM for the ORMDL3 gene? What are the other genes with above zero FPKM values?

The FPKM for ORMDL3 was 137590. The other genes with FPKM values were ZPBP2, GSDMB, and	PSMD3.




