---
title: "Lab15"
author: "Max Wang"
format: pdf
---
## Background

**Git** is an installable version control program which stores information about version history within a document. This is useful in tandem with online git repositories like **github**, allowing for elegant offline collaboration and removes the necessity of servers to host projects. Through the github api, snapshots can be pushed directly through programming languages like R using user tokens. 

The core steps to use git:

1. Imitate a git repository 

>in R studio you can click the radio button on project creation or type `git init` in the comand line

2. Add to the project, adding files, modifying files, etc

3. Add important files to the **stage** which holds files to be committed

>in R studio you chcek "staged" checkmark in the git tab next to the enviroment. in the command line you can type `git add filename`

4. Commit files into the git repository

> in R studio, hit the "commit" buttion in the git tab. in the comand line type `git commit - m "messege"`

5. With github you can push commits into the online git repositories. 

## Using Github API

Using the `usethis` library we can create a Personal Access Token (PAT)

```{r}
library(usethis)
#create_github_token() directs to a web page to create a PAT
```



Creating a github personal access token using the `gitcreds` library 

```{r}
library(gitcreds)
# gitcreds_set()
```

