---
title: "Lab06"
author: "Max Wang (PID A18629546)"
format: pdf
toc: true
---

## Background 
Functions act to condense code into simple, repeatable calls to lines of code which makes up the heart of R. 

All functions in R have 3 components:

1. A **name** to call functions 
2. One or more **arguments** that are fed into the function separated by commas 
3. The **body** of the function, lines of code that describe what the function does 

## First exploration with functions in R

Writing a function that adds 1 to arg x 

```{r}
add <- function(x) {
  x + 1 
}

add(2)
```

Imputing a different object types

```{r}
add(c(100,200,300))
```


Modified function to add arg x to arg y

```{r}
add <- function(x,y) {
  x + y
}

add(3,5)
```

Creating default arguments 

```{r}
add <- function(x = 0,y = 0) {
  x + y
}

add(5)
add(y = 5)
```
**N.B.** Input arguments can be either required or optional based on if they have a default value set. This allows functions to select which arguments are necessary to be input and which ones are not while automatically assigning values to unspecified arguments. 

## Second function 

All functions in R have the same basic format:

```
name <- function(arg, arg) {
  body
}
```

The `sample()` function in R takes a random sample from a vector with n of size `size`. Additionally, the sample function takes arguments `replace` and `prob` which define if the sample takes with or without replacement and the probability for each value described in a vector. 

```{r}
  sample(1:50, 20, replace = T)
  #outputs a random sample of n = 20 from a vector of integers 1 to 50 with replacement
```

>Q. Return 12 numbers picked randomly from the input 1:10

```{r}
  sample(1:10, 12, replace = T)
  #requires replace = T since size > length(x)
```
>Q. Write the code to generate a 12 nt long DNA sequence.

```{r}
seq <- sample(c("A", "T", "G", "C"), 12, T)
cat(seq, sep = "")
```

>Q. Write a first version function called `generate_dna()` that generates a user specified length `n` random DNA sequence 

>Q. Modify the output to return a FASTA like sequence rather than a vector output

```{r}
generate_dna <- function(n){
  seq = sample(c("A", "T", "G", "C"), n, T)
  return(paste(seq, collapse = ""))
  ## similar output as using cat(seq, sep = "")
}
```

```{r}
generate_dna(25)
```

Adding the option to return as a FASTA format or a vectorized output 

```{r}
generate_dna <- function(n, fasta = T){
  seq = sample(c("A", "T", "G", "C"), n, T)
  
  if (fasta){
    return(paste(seq, collapse = ""))
  }
  
  else{
    return(seq)
  }
}
```

```{r}
generate_dna(25) #FASTA format
generate_dna(25,F) #vector format
```

## Third function
>Q. Write a function called `generate_protein()` that generates a user specified length proteins equeince in FASTA like format

```{r}
generate_protein <- function(n){
  aas <- c("A", "G", "V", "L", "I", "P", "F", "Y", "W", "S", "T", "C", "M", "D", "E", "K", "R", "H", "N", "Q")
  seq <- sample(aas, n, T)
  seq <- paste(seq, collapse = "")
  return(seq)
}
```

```{r}
generate_protein(15)
```



>Q. Use your new `generate_protein()` function to generate sequences betweene length 6 and 12 amino acids in length and check if anyh of these are unique in nature (using NR in NCBI database)? 

```{r}
#using a for loop
for(i in 6:12){
  cat(">", i, "\n", sep = "")
  cat(generate_protein(i), "\n")
}
```

